---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dev = "png",
  dpi = 150,
  fig.height = 6,
  fig.width = 9,
  out.width = "100%"
)
```

# ffc

> **F**unctional **F**ore**C**asting

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/ffc)](https://CRAN.R-project.org/package=ffc)
[![R-CMD-check](https://github.com/nicholasjclark/ffc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/nicholasjclark/ffc/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/nicholasjclark/ffc/graph/badge.svg)](https://app.codecov.io/gh/nicholasjclark/ffc)


The goal of the `ffc` 📦 is to forecast complex, time-changing functional relationships by integrating Generalized Additive Models with dynamic factor functional basis expansions. 

**Key benefits:**

- Model functional responses that change shape over time (not just magnitude)  
- Forecast entire curves into the future, not just single values  
- Handle complex multivariate time series with functional structure  
- Seamless integration with the powerful   [`mgcv`](https://cran.r-project.org/package=mgcv) and [`fable`](https://fable.tidyverts.org/) ecosystems

The package introduces dynamic functional predictors using the new `fts()` term, which decomposes functional time series into time-varying basis coefficients that can be forecasted using either independent time series models from the `fable` package or with efficient dynamic factor models using precompiled Stan models.

## Installation

You can install the development version of `ffc` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("nicholasjclark/ffc")
```

## Quick Start

```{r include = FALSE}
library(ffc)
```


```{r eval=FALSE}
library(ffc)

# Fit a model with time-varying coefficients
mod <- ffc_gam(
  response ~ fts(predictor, time_k = 10),  
  data = your_data,
  time = "time_column",
  family = gaussian()
)

# Forecast the functional coefficients
fc <- forecast(mod, newdata = future_data, model = "ARDF")
```

## Examples

### Tourism Forecasting with `fabletools` Integration
```{r, message=FALSE, warning=FALSE}
library(fable)
library(tsibble)
library(tidyverse)
library(ggplot2); theme_set(theme_bw(base_size = 12))
```

Our aim here is to forecast the number of domestic visitors to Melbourne, Australia. The data can be found in the `tsibble::tourism` data set. For now we need to explicitly add the `quarter` and `time` variables to the data, but in future this will be done automatically for seamless integration with the [`tsibbleverse`](https://tsibble.tidyverts.org/)

```{r tourism-data}
tourism_melb <- tourism |>
  filter(
    Region == "Melbourne",
    Purpose == "Visiting"
  ) |>
  mutate(
    quarter = lubridate::quarter(Quarter),
    time = dplyr::row_number()
  )
tourism_melb
```

Split into training and testing folds. We wil aim to forecast the last 5 quarters of the data
```{r train-test-split}
train <- tourism_melb |>
  dplyr::slice_head(n = 75)

test <- tourism_melb |>
  dplyr::slice_tail(n = 5)
```

Now fit an `ffc_gam`. We use time-varying level and time-varying seasonality components, together with a Tweedie observation model (because our outcome, `Trips`, consists of non-negative real values). This model is simpler so we use the `'gam'` engine for fitting:
```{r fit-tourism-model, fig.show='hide'}
mod <- ffc_gam(
  Trips ~
    # Use mean_only = TRUE to model a time-varying mean
    fts(
      time,
      mean_only = TRUE,
      time_k = 50,
      time_m = 1
    ) +
    # Time-varying seasonality
    fts(
      quarter,
      k = 4,
      time_k = 15,
      time_m = 1
    ),
  time = "time",
  data = train,
  family = tw(),
  engine = "gam"
)
```

The `autoplot()` method is handy for viewing the time-varying basis coefficients from `ffc_gam()` models. Here we draw 10 realisations from the estimated coefficient distributions and plot them.

```{r tourism-coefficients0, fig.alt="autplot() plots showing time-varying coefficients for the tourism model, illustrating how basis coefficients have been estimated to vary through time."}
fts_coefs(mod, times = 10, summary = FALSE) |>
  autoplot()
```

We can also draw the time-varying basis coefficients using support from the [`gratia`](https://gavinsimpson.github.io/gratia/) package, which has helpful functions for plotting smooth effects:

```{r tourism-coefficients, fig.alt="Smooth plots showing time-varying coefficients for the tourism model. The upper panel shows the time-varying level component while the lower panel displays seasonal coefficients that change over time, demonstrating how seasonality patterns evolve."}
gratia::draw(mod)
```

These plots show the time-varying *coefficients* for a set of basis functions. We have one function representing the mean of the series (essentially a constant) as well as three basis functions representing the quarterly seasonality. Each of these has a coefficient that can change through time, allowing the entire functional series to change shape over time. We can compute forecast distribution by fitting the basis coefficient forecast models in parallel (which is automatically supported within the `fable` package). Here we fit independent exponential smoothing models to each coefficient time series
```{r tourism-forecast, fig.show='hide'}
fc <- forecast(
  object = mod,
  newdata = test,
  model = "ETS",
  summary = FALSE
)
```

We can also convert resulting forecasts to a `fable` object for automatic plotting and/or scoring of forecasts
```{r build-fable, warning=FALSE}
# Using the new as_fable method for seamless conversion
fc_ffc <- as_fable(mod, newdata = test, forecasts = fc)
fc_ffc
```

Leverage the fabletools ecosystem for forecast analysis
```{r fabletools-analysis, warning=FALSE}
# Calculate accuracy metrics
accuracy(fc_ffc, test)

# Generate prediction intervals  
fc_intervals <- hilo(fc_ffc, level = c(80, 95))
fc_intervals

# Distribution summaries
fc_summary <- fc_ffc |>
  summarise(
    mean_forecast = mean(.dist),
    median_forecast = median(.dist),
    q25 = quantile(.dist, 0.25), 
    q75 = quantile(.dist, 0.75)
  )
fc_summary
```

Next we can explore how to compare forecasts from `ffc` models to traditional time series models by again leveraging the simplicity and power of the `fable` ecosystem

```{r model-comparison, warning=FALSE, fig.alt="Comprehensive model comparison showing FFC functional forecasting versus traditional ARIMA and ETS models for Melbourne tourism data. The plot demonstrates forecast performance with MAPE values in titles, revealing the added value of functional forecasting approaches."}
# Generate FFC forecasts with different models
fc_ffc_arima <- as_fable(mod, newdata = test, model = "ARIMA")
fc_ffc_ets <- as_fable(mod, newdata = test, model = "ETS")

# Generate traditional model forecasts
fc_traditional <- train |>
  model(
    ARIMA = ARIMA(Trips),
    ETS = ETS(Trips)
  ) |>
  forecast(h = 5)

# Calculate accuracy for all models
acc_ffc_arima <- accuracy(fc_ffc_arima, test)
acc_ffc_ets <- accuracy(fc_ffc_ets, test)
acc_traditional <- accuracy(fc_traditional, test)

# Extract MAPE values for titles
mape_ffc_arima <- round(acc_ffc_arima$MAPE, 1)
mape_ffc_ets <- round(acc_ffc_ets$MAPE, 1)
mape_arima <- round(acc_traditional$MAPE[acc_traditional$.model == "ARIMA"], 1)
mape_ets <- round(acc_traditional$MAPE[acc_traditional$.model == "ETS"], 1)

# Create comparison plots
library(patchwork)

p1 <- autoplot(fc_ffc_arima, train) + 
  geom_line(data = test, aes(y = Trips), color = "black") +
  ggtitle(paste0("FFC with ARIMA (MAPE: ", mape_ffc_arima, "%)"))

p2 <- autoplot(fc_ffc_ets, train) + 
  geom_line(data = test, aes(y = Trips), color = "black") +
  ggtitle(paste0("FFC with ETS (MAPE: ", mape_ffc_ets, "%)"))

p3 <- autoplot(filter(fc_traditional, .model == "ARIMA"), train) +
  geom_line(data = test, aes(y = Trips), color = "black") +
  ggtitle(paste0("Traditional ARIMA (MAPE: ", mape_arima, "%)"))

p4 <- autoplot(filter(fc_traditional, .model == "ETS"), train) +
  geom_line(data = test, aes(y = Trips), color = "black") +
  ggtitle(paste0("Traditional ETS (MAPE: ", mape_ets, "%)"))

(p1 | p2) / (p3 | p4)
```

These plots illustrate how the `ffc` models outperform traditional forecasting models for this forecasting experiment by thinking about this as a functional time series problem.

## Getting help
If you encounter a clear bug, please file an issue with a minimal reproducible example on [GitHub](https://github.com/nicholasjclark/ffc/issues)

## Contributing
Contributions are very welcome, but please see our [Code of Conduct](https://github.com/nicholasjclark/ffc/blob/main/.github/CODE_OF_CONDUCT.md) when you are considering changes that you would like to make.

## License
The `ffc` project is licensed under an `MIT` open source license
