---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dev = "png",
  dpi = 150,
  fig.height = 6,
  fig.width = 9,
  out.width = "100%"
)
```

# ffc

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/ffc)](https://CRAN.R-project.org/package=ffc)
[![R-CMD-check](https://github.com/nicholasjclark/ffc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/nicholasjclark/ffc/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/nicholasjclark/ffc/graph/badge.svg)](https://app.codecov.io/gh/nicholasjclark/ffc)
<!-- badges: end -->

# ffc

> **F**unctional **F**ore**C**asting

The goal of the `ffc` 📦 is to perform functional regression using Generalized Additive Models (GAMs). The package integrates with the extremely flexible [`mgcv`](https://cran.r-project.org/package=mgcv){target="_blank"} package to enable functional responses to be modelled and predicted using a broad range of predictor effects. Key among these types of predictors are *dynamic functional predictors* using a new `fts()` term, which sets up functional predictors whose coefficients are modelled as time-varying. These time-varying coefficients can then be forecasted ahead using a variety of efficient forecasting algorithms, providing unmatched flexibility to model and predict how functional responses change over time.

## Installation

You can install the development version of ffc from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("nicholasjclark/ffc")
```

## A brief example
Load the in-built Queensland Mortality data, which contains the number of deaths per age category over time in the state of Queensland, Australia
```{r, message=FALSE, warning=FALSE}
library(ffc)
library(ggplot2)
theme_set(theme_bw())
data("qld_mortality")
head(qld_mortality, 15)
```

Visualise the observed mortality curves over time using the log10 scale
```{r, warning=FALSE}
ggplot(
  data = qld_mortality,
  aes(
    x = age,
    y = deaths / population,
    group = year,
    colour = year
  )
) +
  geom_line() +
  facet_wrap(~sex) +
  scale_colour_viridis_c() +
  labs(y = "Observed log(Mortality)") +
  scale_y_log10()
```

Fit a model to estimate how the log(mortality) curve changed over time using `deaths` as the outcome and using a time-varying function of `age` as the primary predictor. Using `fts()`, we model the age-death functions hierarchically using thin plate basis functions whose coefficients are allowed to vary over time, where `time = 'year'`. The hierarchical formulation allows a shared time-varying level to be modelled, along with deviations around that time-varying level for each sex. We use the `bam()` engine (as opposed to `gam()`) for parameter estimation, given the large size of the dataset. In future, other engines such as `brm()` and `mvgam()`, will be made available for full luxury Bayesian inference.
```{r}
mod <- ffc_gam(
  deaths ~
    offset(log(population)) +
    sex +
    # Use mean_only = TRUE to model a time-varying level
    fts(
      year,
      mean_only = TRUE,
      bs = "tp",
      time_k = 35,
      time_m = 1
    ) +
    # Model the male/female deviations around the shared mean
    fts(
      age,
      by = sex,
      bs = "tp",
      time_k = 15,
      time_m = 1
    ),
  time = "year",
  data = qld_mortality,
  family = poisson(),
  engine = "bam"
)
```

Inspect the model summary; notice in the `Formula` slot how the basis functions are modelled as `by` variables within independent smooths of `year` that share their smoothing parameters
```{r}
summary(mod)
```

View predicted functional curves using a fixed offset (where `population = 1`), which allows us to calculate a standardized rate of mortality
```{r}
newdat <- qld_mortality
newdat$population <- 1
newdat$preds <- predict(
  mod,
  newdata = newdat,
  type = "response"
)

ggplot(
  data = newdat,
  aes(
    x = age,
    y = preds,
    group = year,
    colour = year
  )
) +
  geom_line() +
  facet_wrap(~sex) +
  scale_colour_viridis_c() +
  labs(y = "Expected log10(Mortality)") +
  scale_y_log10()
```

Using support from the `marginaleffects` 📦, we can make easily predict changes in mortality rate for specific age groups. For example, here is the expected decline in mortality rate for 17 year-olds in Queensland over the study period
```{r warning=FALSE}
library(marginaleffects)
plot_predictions(
  mod,
  by = c("year", "sex"),
  newdata = datagrid(
    age = 17,
    year = unique,
    sex = unique,
    population = 1
  ),
  type = "response"
) +
  labs(
    x = "Year",
    y = "Expected log10(Mortality)"
  ) +
  scale_y_log10()
```

And here are the slopes of this change
```{r, warning=FALSE}
plot_slopes(
  mod,
  variables = "year",
  by = c("year", "sex"),
  newdata = datagrid(
    age = 17,
    year = unique,
    sex = unique,
    population = 1
  ),
  type = "response"
) +
  labs(
    x = "Year",
    y = "1st derivative of mortality rate change"
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed"
  )
```

The time-varying coefficients can be extracted into a `tidy` format using `fts_coefs()`, which will facilitate the use of time series models to enable efficient forecasting of the entire curve into the future. Using `summary = FALSE` will return draws of each coefficient time series from the model's empirical Bayesian posterior distribution (you can control the number of draws that are returned using the `times` argument):
```{r}
functional_coefs <- fts_coefs(
  mod,
  summary = FALSE,
  times = 10
)
functional_coefs
```

The basis function coefficient time series can be plotted using `autoplot()`
```{r}
autoplot(functional_coefs)
```

Clearly there is a lot of structure and dependence here, suggesting that a dynamic factor model fitted to these coefficient time series would be valuable for creating functional forecasts. But for now we can apply any model from the `fable` 📦 to these replicate time series and generate future forecast realisations, which can be summarised to approximate the full uncertainty in our coefficient forecast distributions. Again here you can control the number of forecast paths that are simulated from the underlying time series models using the `times` argument
```{r, warning=FALSE}
functional_fc <- forecast(
  object = functional_coefs,
  h = 5,
  times = 5
)
functional_fc
```

Another example showing univariate time series modelling and forecasting
```{r, message=FALSE, warning=FALSE}
library(fable)
library(tsibble)
library(dplyr)
```

Our aim here is to forecast the number of domestic holiday travellers to Melbourne, Australia. The data can be found in the `tsibble::tourism` data set. For now we need to explicitly add the `quarter` and `time` variables to the data, but in future this will be done automatically for seamless integration with the `tsibbleverse`
```{r}
tourism_melb <- tourism %>%
  filter(
    Region == "Melbourne",
    Purpose == "Holiday"
  ) %>%
  mutate(
    quarter = lubridate::quarter(Quarter),
    time = dplyr::row_number()
  )
tourism_melb
```

Split into training and testing folds. We wil aim to forecast the last 5 quarters of the data
```{r}
train <- tourism_melb %>%
  dplyr::slice_head(n = 75)

test <- tourism_melb %>%
  dplyr::slice_tail(n = 5)
```

Now fit an `ffc_gam`. We use time-varying level and time-varying seasonality components, together with a Tweedie observation model (because our outcome, `Trips`, consists of non-negative real values). This model is simpler so we use the `'gam'` engine for fitting:
```{r}
mod <- ffc_gam(
  Trips ~
    # Use mean_only = TRUE to model a time-varying mean
    fts(
      time,
      mean_only = TRUE,
      time_k = 50,
      time_m = 1
    ) +
    # Time-varying seasonality
    fts(
      quarter,
      bs = "tp",
      k = 4,
      time_k = 15,
      time_m = 1
    ),
  time = "time",
  data = train,
  family = tw(),
  engine = "gam"
)
```

Draw the time-varying basis coefficients, this time using `gratia`
```{r}
gratia::draw(mod)
```

Compute forecast distribution by fitting the basis coefficient time series models in parallel (which is automatically supported within the `fable` package). Here we fit independent exponential smoothing models to each coefficient time series
```{r}
fc <- forecast(
  object = mod,
  newdata = test,
  model = "ETS",
  summary = FALSE
)
```

Convert resulting forecasts to a `fable` object for automatic plotting and/or scoring of forecasts
```{r, warning=FALSE}
test[["value"]] <- fc

fc_ffc <- fabletools:::build_fable(
  test,
  response = "Trips",
  distribution = "value"
)
```

Plot the forecast distribution along with the observed data
```{r}
fc_ffc %>%
  autoplot(train) +
  geom_line(
    data = test,
    aes(y = Trips)
  ) +
  ggtitle("FFC forecast")
```

Compare to forecasts from automatic ARIMA and ETS models, which are simpler to code and of course a bit faster
```{r}
train %>%
  model(
    arima = ARIMA(Trips)
  ) %>%
  forecast(
    h = 5
  ) %>%
  autoplot(train) +
  geom_line(
    data = test,
    aes(y = Trips)
  ) +
  ggtitle("ARIMA forecast")
```

```{r}
train %>%
  model(
    ets = ETS(Trips)
  ) %>%
  forecast(
    h = 5
  ) %>%
  autoplot(train) +
  geom_line(
    data = test,
    aes(y = Trips)
  ) +
  ggtitle("ETS forecast")
```

## Getting help
If you encounter a clear bug, please file an issue with a minimal reproducible example on [GitHub](https://github.com/nicholasjclark/ffc/issues)

## Contributing
Contributions are very welcome, but please see our [Code of Conduct](https://github.com/nicholasjclark/ffc/blob/main/.github/CODE_OF_CONDUCT.md) when you are considering changes that you would like to make.

## License
The `ffc` project is licensed under an `MIT` open source license
